Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2019-09-29 astronum_2019_Castro/astronum-castro-2019.tex
start of the AstroNum 2019 paper using last year's as the template

Flame spreading was first modeled in a series of calculations
employing an algorithm from atmospheric science, where vertical
hydrostatic equilibrium is enforced and wide-aspect ratio zones give a
horizontal CFL number that is large enough to obtain long timescale
evolution~\cite{cavecchi:2012}.  These calculations display key
features of the propagation mechanism, namely a balance between
hydrodynamics and the pure flame physics. The flame proceeds mainly
via conduction, thus being a deflagration. The flame front, however,
is not vertical but inclined, the angle with the horizontal being very
small: $\theta \sim H/R_{\rm{Ro}} \sim 10^{-3}$. Here $H$ is the scale
height and $R_{\rm{Ro}}$ is the Rossby radius. It is the
hydrodynamical balance between the Coriolis force, gravity and
pressure gradients that determines the Rossby radius and therefore the
inclination angle of the flame front. The extended surface of the
flame front is what leads to propagation speeds of order $10^5 $~km/s,
in good agreement with observations. Due to its dependence on the
Coriolis force, the flame speed changes across the surface of the NS
(being faster at the equator), leading to potentially detectable
effects if the rise of the burst is observed with enough resolution
\cite{art-2015-cavecchi-etal}.  We note that no calculations have
considered the effects of the flame interacting with preexisting
turbulence ahead of the flame from simmering that may be occurring
there, as we characterized in \cite{xrb3}.  Flame-turbulence
interactions have long been seen as a means to accelerate a flame in
Type Ia supernovae, so future studies should look at turbulence in the
layer.  Capturing turbulence, of course, requires high resolution, so
the ideas described here will be needed.

@@

Machine architectures are evolving, with heterogeneous architectures
becoming more common.  Our recent development in \castro\ has focused
on moving the hydrodynamics to GPUs~\cite{astronum:2017}.  The XRB
problem is an ideal case for this, since hydrodynamics with constant
gravity (for plane-parallel domains) ports in a straightforward manner
to GPUs.  The current GPU version of the \castro\ hydrodynamics solver
is about $10\times$ faster using GPUs on a node than the entire node
of CPU cores (comparing 4 NVIDIA Volta GPUs to 20 IBM Power9 cores).
We see similar speed ups with reaction networks, with these GPU ports
under active development.  This means a GPU calculation could lower
the above costs by a factor of 10.  GPU offloading of the reaction
networks will also allow us to explore larger networks and more
detailed nuclear physics.

>>>>

Flame spreading was first modeled in a series of calculations
employing an algorithm from atmospheric science, where vertical
hydrostatic equilibrium is enforced and wide-aspect ratio zones give a
horizontal CFL number that is large enough to obtain long timescale
evolution~\cite{cavecchi:2012}.  These calculations display key
features of the propagation mechanism, namely a balance between
hydrodynamics and the pure flame physics. The flame proceeds mainly
via conduction, thus being a deflagration. The flame front, however,
is not vertical but inclined, the angle with the horizontal being very
small: $\theta \sim H/R_{\rm{Ro}} \sim 10^{-3}$. Here $H$ is the scale
height and $R_{\rm{Ro}}$ is the Rossby radius. It is the
hydrodynamical balance between the Coriolis force, gravity and
pressure gradients that determines the Rossby radius and therefore the
inclination angle of the flame front. The extended surface of the
flame front is what leads to propagation speeds of order $10^5 $~km/s,
in good agreement with observations. Due to its dependence on the
Coriolis force, the flame speed changes across the surface of the NS
(being faster at the equator), leading to potentially detectable
effects if the rise of the burst is observed with enough resolution
\cite{art-2015-cavecchi-etal}.  We note that there may be preexisting
turbulence ahead of the flame from simmering occurring
there, and no calculations have considered the effects of the flame
interacting with this turbulence, as we characterized in \cite{xrb3}.
Flame-turbulence interactions have long been seen as a means to
accelerate a flame in Type Ia supernovae, so future studies should
look at turbulence in the layer.  Capturing turbulence, of course,
requires high resolution, so the ideas described here will be needed.

@@

Machine architectures are evolving, with heterogeneous architectures
becoming more common.  Our recent development in \castro\ has focused
on moving the hydrodynamics to GPUs~\cite{astronum:2017}.  The XRB
problem is an ideal case for this, since hydrodynamics with constant
gravity (for plane-parallel domains) ports in a straightforward manner
to GPUs.  The current GPU version of the \castro\ hydrodynamics solver
is about $10\times$ faster using GPUs on a node than the entire node
of CPU cores (comparing 4 NVIDIA Volta GPUs to a dual-socket IBM Power9 CPU using 40 cores).
We see similar speed ups with reaction networks, with these GPU ports
under active development.  This means a GPU calculation could lower
the above costs by a factor of 10.  GPU offloading of the reaction
networks will also allow us to explore larger networks and more
detailed nuclear physics.
